---
title: "P8105 HW3"
author: "Lucia Wang (lw3061)"
date: "due 2023-10-14"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(p8105.datasets)
knitr::opts_chunk$set(echo = TRUE)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

## Problem 1 = Instacart
```{r}
data("instacart")
```

```{r, include=FALSE}
# data exploration
nrow(instacart)

instacart |> 
  group_by(aisle) |> 
  summarize(n_aisles = n()) |> 
  nrow()

instacart_sort = instacart |>
  group_by(aisle) |>
  summarize(n_aisles = n()) |>
  mutate(
    rank_aisles = rank(n_aisles)
  ) |> 
  arrange(rank_aisles)

instacart_aisles = instacart |>
  group_by(aisle) |> 
  summarize(n_aisles = n()) |>
  filter(n_aisles > 10000) |>
  mutate(
    rank_aisles = rank(n_aisles)
  ) |>
  arrange(rank_aisles)

```
This dataset shows grocery store items including the aisle number and what product it is. It has information on if people added it to the cart or reordered it, as well as what time of day it was ordered.

1. # of aisles and aisles with most items ordered

There are `r nrow(instacart)` items in the data, across `r nrow(summarize(group_by(instacart, aisle), n_aisles=n()))` aisles. The top 20 aisles from which the most items are ordered is `r select(filter(instacart_sort, rank_aisles <= 20), aisle, rank_aisles)`. 

2. Plot with # of items ordered per aisle, only for aisles > 10000 items ordered.

```{r}
instacart_aisles |>
  ggplot(aes(x=rank_aisles, y=n_aisles, color=aisle)) +
  geom_point() +
  labs(
    x="Rank of aisle from most to least items ordered",
    y="Number of items"
  ) +
  theme(legend.position="none") +
  geom_text(aes(label=aisle), size=1, hjust=1, vjust=1, angle=90)
```




## Problem 2 = BRFSS
First, clean and format the data.
```{r}
data("brfss_smart2010")

brfss = brfss_smart2010 |>
  janitor::clean_names() |>
  filter(topic=="Overall Health", response %in% c("Excellent","Very good","Good","Fair","Poor")) |>
  mutate(
    response = factor(response, levels=c("Poor", "Fair", "Good", "Very good", "Excellent"))
  )
```

*1. In 2002, which states were observed at 7 or more locations? What about in 2010?*
```{r}
states_02 = brfss |>
  filter(year==2002) |>
  separate(locationdesc, into =c("state","county"), sep =5) |>
  select(-state) |>
  group_by(locationabbr) |>
  summarize(n_obs = n()) |>
  mutate(counties_per_state = n_obs/5)

```

There were `r nrow(filter(states_02, counties_per_state >= 7))` states observed at 7 or more locations. They were `r pull(filter(states_02, counties_per_state >= 7),locationabbr)`.

```{r}
states_10 = brfss |>
  filter(year==2010) |>
  separate(locationdesc, into =c("state","county"), sep =5) |>
  select(-state) |>
  group_by(locationabbr) |>
  summarize(n_obs = n()) |>
  mutate(counties_per_state = n_obs/5)
```

There were `r nrow(filter(states_10, counties_per_state >= 7))` states observed at 7 or more locations. They were `r pull(filter(states_10, counties_per_state >= 7),locationabbr)`

*2. Filter to only `Excellent` responses and make a spaghetti plot with a newly calculated average value variable.*
```{r}
excellent = brfss |> 
  filter(response=="Excellent") |>
  separate(locationdesc, into =c("state","county"), sep =5) |>
  select(-state) |>
  group_by(year,locationabbr) |>
  summarize(avg_dataval = mean(data_value))

excellent_plot = excellent |> 
  ggplot(aes(x=year, y=avg_dataval, group=locationabbr)) + 
  geom_line(aes(color=locationabbr))

excellent_plot
```

The average data value has fluctuated over the years, with a very general trend of decreasing between 2002 and 2010. Most fall in the range between 17.5 to 27.5, but a few outlier states have average values closer to 10 or 30. 

*3. Two-panel plot for `data_value` distribution in NY for 2006 vs 2010.*
```{r}
ny_state = brfss |>
  filter(year==2006|year==2010,locationabbr=="NY") |>
  separate(locationdesc, into =c("state","county"), sep =5) |>
  select(-state) |>
  group_by(county)

ny_plot = ny_state |>
  ggplot(aes(x=data_value, y=response)) +
  geom_point(aes(color=county)) +
  facet_grid(.~year)

ny_plot
```

The highest `data_value` values were for the responses of `Very good` and `Good`. The scores tended to increase between 2006 and 2010 and the ranges of scores per response category also widened. 


## Problem 3 = NHANES
*1. Load, tidy, merge, and organize the data.*
```{r}
demographics = read_csv("nhanes_covar.csv", skip=4) |>
  janitor::clean_names() |>
  mutate(
    sex = case_match(sex,
    1 ~ "Male",
    2 ~ "Female"),
    education = case_match(education,
    1 ~ "Less than high school",
    2 ~ "High school equivalent",
    3 ~ "More than high school"),
    education = factor(education, levels=c("Less than high school", "High school equivalent", "More than high school")),
    seqn = as.character(seqn)
  )  |>
  filter(age >= 21) |>
  drop_na()

accelerometer = read_csv("nhanes_accel.csv") |>
  janitor::clean_names() |>
  mutate(
    seqn = as.character(seqn)
  )

demo_accel = left_join(demographics, accelerometer, by="seqn")
```

The full dataset of demographic variables and accelerometer data has `r nrow(demo_accel)` observations.

*2. Produce a table for sex vs education and make a visualization of the age distributions for each education category.*
```{r}
education = demo_accel |>
  group_by(sex, education) |>
  summarize(n_obs = n())

education |> knitr::kable()
```

For both females and males, the number of observations is greatest in the "more than high school" education category. There are also more males overall. 

```{r}
demo_accel |>
  ggplot(aes(x=age, fill=sex)) +
  geom_histogram(bins=10) +
  facet_wrap(sex~education)
```

The age distributions for less than high school and high school equivalent look similar across sex. 

Females who had more than high school education were younger on average as shown by the skewed right age distribution. While males were also younger on average in the more than high school education category, there seems to be more of a bi-modal distribution as most males were either below 50 years or above 60 years old.

*3. Aggregate across minutes and plot age vs total activity by sex and education.*
```{r}
demo_accel |>
  mutate(
    total_activity = rowSums(across(c(min1:min1440)))
  ) |>
  select(-c(min1:min1440)) |>
  group_by(sex) |>
  ggplot(aes(x=age, y=total_activity, color=sex)) +
  geom_point() +
  geom_smooth(se=FALSE) +
  facet_wrap(.~education)
```

For all education categories, total activity decreased with age. 

In the less than high school category, males on average had more activity than females, surprisingly more at older ages. There was also a wider range of total activity for this education group.

At higher levels of education, females had more total activity than males. High school equivalent females were especially more active than males for ages around 30-50. 

For people with more than high school education, there was less activity overall, and males and females were more similar to each other. However, the trend lines for this education category did not really capture the almost "random"-looking dispersion of age and activity distribution. 

*4. Make a 3-panel plot for the 24-hour activity time by education level and use color to indicate sex.*
```{r}
activity24 = demo_accel |>
  pivot_longer(
    min1:min1440,
    names_to = "min",
    values_to = "activity"
  ) |>
  separate(min, into=c("minname", "minnum"), sep=3) |>
  mutate(
  minnum = as.numeric(minnum)
  ) 

activity24 |>
  ggplot(aes(x=minnum, y=activity, color=sex)) + 
  geom_point(alpha=0.25, size=0.1) +
  geom_smooth() +
  facet_wrap(.~education) +
  labs(
    x = "Hour of day", y="Activity", title ="Activity over 24 hours by sex and education category"
  )
```

Most participants had activity levels between 0-50 when looking at their activity data by hour. 

As expected, the most activity occurred during the day, with a dip in the early hours of the day. The peaks of all 3 panels was around midday.

Females tended to have more activity than males when looking at trend lines. 

Education did not seem to have an effect on amounts of activity in general, but when the data is filtered to people with `activity` < 50 per minute, those in the less than high school category trended with slightly more activity than the other two education categories. 
```{r}
activity24 |>
  filter(activity < 50) |>
  ggplot(aes(x=minnum, y=activity, color=sex)) + 
  geom_point(alpha=0.25) +
  geom_smooth() +
  facet_wrap(.~education) +
  labs(
    x = "Hour of day", y="Activity", title ="Activity over 24 hours, with values >50 excluded"
  )
```


